---
title: "README"
author: "Ezinne Ndukwe"
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# R/dpdquery

## Overview

The DPD Query package (dpdquery) is a suite of functions designed to
query Health Canada's Drug Product Database (DPD) through component
APIs. The DPD is managed by Health Canada and provides information on
drugs approved and authorized for sale in Canada.

This package is intended to assist with queries of the DPD. The results
are returned in an R-based structured data frame that can be exported to
external files.

I did not make this package as an employee of Health Canada, nor did I
receive any form of payment for making this package. For more
information regarding the DPD and the information provided, please
consult Health Canada resources and contact their team.

- [Drug Product Database: Access the
  database](https://www.canada.ca/en/health-canada/services/drugs-health-products/drug-products/drug-product-database.html)

- [Drug Product Database (DPD) API
  Guide](https://health-products.canada.ca/api/documentation/dpd-documentation-en.html)

  ------------------------------------------------------------------------

## Installation

```{r message=FALSE, warning=FALSE}
# Install development version from GitHub
# install.packages("devtools")
devtools::install_github("ezyndukwe/dpdquery", upgrade = 'never')
library(dpdquery)
```

## Purpose

As a researcher in pharmcoepidemiology, I realized that I could not find
a way to export the results of the Health Canada Drug Product Database
online query. I made this package to integrate the results of the online
query into my R-based environment. The package returns results in data
frame which can then be written into tabular data files and used in many
programming software.

This package is designed to query the DPD in a replicable, verifiable
and efficient process. Using the `query_dpd_by_activeIngredient()`
function in this package, I was able to search the entire DPD for 35
diabetic active ingredients and retrieve information on 990 drug
products in a structured table in just 13 minutes. I was able to review
the output results and confirm that the active ingredient terms that did
not return any drug products were not available in Canada (and not
typos).

For individuals who wish to retrieve all the same information found in
the DPD online query but using an R-based tool, this package is for you!
You can use the functions in the package that being with
"`query_dpd_by_`" to access multiple component APIs of the DPD and run a
comprehensive query that replicates the results of the online DPD query.
You can query the DPD by active ingredient, brand name (product name),
DIN or drug code. You can also use functions that start with "`search_`"
to access the component APIs.

------------------------------------------------------------------------

## Vignette

For an examples of a workflow using the dpdquery, please see the
vignette I prepared which can be accessed with the code
`vignette(package = 'dpdquery')`.
