---
title: "DPD Query Vignette - Get started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DPD Query Vignette - Get started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r eval=FALSE, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Suggested workflow

```{r setup, echo=TRUE, message=FALSE, warning=FALSE, setecho=TRUE}
library(dpdquery)
```

This document shows a suggested for researchers or organizations who use the Drug Product Database from Health Canada. The provided example will be on Ozempic and generic versions of semaglutide products.

## Scoping for products with a given brand name

On January 4, 2018, Health Canada approved Ozempic (semaglutide injections) for adults with type 2 diabetes mellitus. With that, we know the injection pens are available with different dosage strengths. We can use the `search_dp_api()` function (with the default `find_ais=TRUE`) to search for different dosage strengths that Ozempic comes in. At the time of this search, 6 products have been approved.

```{r query_brand_name, echo=TRUE}
ozempic_search <- search_dp_api(brand_name = 'ozempic')
ozempic_search

```

## Confirming drug status

Of the 6 products that were found, we cannot tell if all of them are still approved in Canada. We can check this using the `search_productstatus_api()` function. We can only query the Product Status API of the DPD using drug product codes (this is different from the drug product's Drug Identification Number (DIN)). We can find the drug product codes of the retrieved Ozempic products in the "drug_code" column of the output data frame of our previous search.

```{r}
ozempic_dc <- ozempic_search$drug_code
ozempic_statuses <- search_productstatus_api(ozempic_dc)
ozempic_statuses
```

From the search, we see that 3 products have "Marketed" status and the other 3 have "Approved" status. We can check [DPD terminology](https://www.canada.ca/en/health-canada/services/drugs-health-products/drug-products/drug-product-database/terminology.html){.uri} page for more information for what each status means.

- **Approved:** refers to an active DIN for a product that has been reviewed and authorized for sale in Canada but has not yet been marketed in Canada.

- **Marketed:** refers to an active DIN for a product that is currently being sold in Canada.

Most of the functions that begin with "`search_`" in this package can only query their component API by drug code. The function `search_ai_api()` can search the Active Ingredient API by drug code(s) or active ingredient(s). The function `search_dp_api()` can search the Drug Product API by drug product code(s), DIN(s) or brand name(s).

## Comprehensive search

### Replicate results of Drug Product Database online query

In practice, when we are doing pharmacological research and projects, we want to know more about drugs than just their dosage strength and status with Health Canada. We may want to know about their active ingredients, dosage forms, class, schedules, routes of administration, classification in the ATC system, and other details. This information can be retrived with the [Drug Product Database online query](https://health-products.canada.ca/dpd-bdpp/).

The reason I made this R package was because I wanted to integrate the results of the online query into my R environment. Unfortunately, after running the query, I could not find a way to export the results in a CSV or Excel file, or any kind of file for that matter. Meaning, I would have had to go through countless pages of the results from my query of 35 active ingredients found in diabetes medications and record the product details manually (I should note that just a search with 15 of these ingredients returns 51 pages with 10 products each). I also realized that I would not know if I had any typos in the terms I would supply to the DPD. Thus, a workflow using the DPD online query is very error-prone, not the most reproducible, and inefficient. Using the tools in this package, I was able to search the entire DPD using the 35 active ingredients in just 13 minutes, retrieving 990 drug products.

For individuals who wish to retrieve all the same information found in the DPD online query but using an R-based tool, this package is for you! You can use the various functions in the package that being with "`query_dpd_by_`" to access multiple component APIs of the DPD and run a comprehensive query that replicates the results of the online DPD query. You can query the DPD by active ingredient, brand name (product name), DIN or drug code.

Continuing off the Ozempic thread, you can use the function `query_dpd_by_brandName()` to search the DPD for information on Ozempic drug products.

```{r quering_DPD_by_brandName, echo=TRUE}
ozempic_query_by_BN <- query_dpd_by_brandName('ozempic')
```

### Search generic products

On April 28, 2026, Health Canada approved a generic semaglutide injection by Dr. Reddy's Laboratories. This was the first generic version of the Ozempic semaglutide injection. When they made the announcement, Health Canada was also in the process of reviewing 8 other submissions for generic semaglutide injections. You can keep up-to-date with the approved semaglutide products with the `query_dpd_by_activeIngredient()` function.

As of May 27, 2026, 4 generic semaglutide injection products (2 by Dr. Reddy's Laboratories and 2 by Apotex) have been authorized and on the market.

```{r quering_DPD_by_active_ingredients, echo=TRUE}
semaglutide_query <- query_dpd_by_activeIngredient('semaglutide', progress_bar = FALSE)
```
